---
title: "make_figures_barcodetrackR"
author: "Ryland Mortlock"
date: "12/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation
Currently, [barcodetrackR](http://github.com/d93espinoza/barcodetrackR) is available at Github and can be downloaded using the devtools package.
```{r installation, eval = FALSE}
devtools::install_github("DunbarLabNIH/barcodetrackR", force = TRUE)
```

## Contributors
The R package and functions were created by Diego A. Espinoza, Ryland D. Mortlock, Samson J. Koelle, and others at Cynthia Dunbar's laboratory at the National Heart, Lung, and Blood Institutes of Health. Issues should be addressed to https://github.com/d93espinoza/barcodetrackR/issues.

# Loading data

## Loading required packages

The [barcodetrackR](http://github.com/d93espinoza/barcodetrackR) package operates on [SummarizedExperiment](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html) objects from the Bioconductor repository. It stores associated colData for each sample in this object as well as any metadata. We load the `barcodetrackR` and `SummarizedExperiment` packages here for our analyses, as well as the `magrittr` package in order to improve legibility of code through using the pipe `%>%` operator.

```{r load packages required, eval = TRUE, warning=FALSE, message=FALSE}
require("magrittr")
require("barcodetrackR")
require("SummarizedExperiment")
require("ggplot2")
require("cowplot")
require("dplyr")
```

## Creating objects with `create_SE`

For this vignette, we will load publically available data from the following papers (these sample datasets are included in the R package):

* [Wu, Chuanfeng, et al. "Clonal expansion and compartmentalized maintenance of rhesus macaque NK cell subsets." Science Immunology (2018)](http://dx.doi.org/10.1126/sciimmunol.aat9781)
* [Koelle, Samson J., et al. "Quantitative stability of hematopoietic stem and progenitor cell clonal output in rhesus macaques receiving transplants." Blood (2017)](https://doi.org/10.1182/blood-2016-07-728691)
* [Belderbos, Mirjam E., et al. "Donor-to-Donor Heterogeneity in the Clonal Dynamics of Transplanted Human Cord Blood Stem Cells in Murine Xenografts." Biology of Blood and Marrow Transplantation (2020)](http://dx.doi.org/10.1016/j.bbmt.2019.08.026)
* [Six, Emmanuelle, et al. "Clonal tracking in gene therapy patients reveals a diversity of human hematopoietic differentiation programs." Blood (2020)](http://doi.org/10.1182/blood.2019002350)
* [Espinoza, Diego A., et al. "Aberrant Clonal Hematopoiesis following Lentiviral Vector Transduction of HSPCs in a Rhesus Macaque." Molecular Therapy 27.6 (2019): 1074-1086.](10.1016/j.ymthe.2019.04.003)
* [Elder A, et al. "Abundant and equipotent founder cells establish and maintain acute lymphoblastic leukaemia." Leukemia (2017)](https://doi.org/10.1038/leu.2017.140)
* [Clarke, Erik L., et al. "T cell dynamics and response of the microbiota after gene therapy to treat X-linked severe combined immunodeficiency." Genome medicine (2018)](https://doi.org/10.1186/s13073-018-0580-z)

```{r load data, eval = TRUE}
system.file("sample_data/WuC_etal/monkey_ZJ31.txt", package = "barcodetrackR") %>%
  read.delim(row.names = 1) -> wu_dataframe

system.file("sample_data/WuC_etal/monkey_ZJ31_metadata.txt", package = "barcodetrackR") %>%
  read.delim() -> wu_metadata

wu_SE <- create_SE(your_data = wu_dataframe,
                   meta_data = wu_metadata,
                   threshold = 0.0001)

system.file("sample_data/KoelleSJ_etal/ZH33_reads.txt", package = "barcodetrackR") %>%
  read.delim(row.names = 1) -> koelle_dataframe

system.file("sample_data/KoelleSJ_etal/ZH33_metadata.txt", package = "barcodetrackR") %>%
  read.delim() -> koelle_metadata

# Fix a typo in Koelle dataset
koelle_metadata["ZH33_14m_T","months"]  <- 14

koelle_SE <- create_SE(your_data = koelle_dataframe,
                       meta_data = koelle_metadata,
                       threshold = 0.0001)

system.file("sample_data/BelderbosME_etal/count_matrix_mouse_C21.txt", package = "barcodetrackR") %>%
  read.delim(row.names = 1) -> belderbos_dataframe

system.file("sample_data/BelderbosME_etal/metadata_mouse_C21.txt", package = "barcodetrackR") %>%
  read.delim() %>%
  set_rownames(value = .$SAMPLENAME) -> belderbos_metadata

belderbos_metadata$celltype <- factor(belderbos_metadata$celltype, levels = c("T", "B", "G", "bulk"))
belderbos_metadata$weeks <- factor(belderbos_metadata$weeks, levels = c("9", "14", "20", "22", "sac"))
belderbos_metadata$organ <- factor(belderbos_metadata$organ, levels = unique(belderbos_metadata$organ))

belderbos_SE <- create_SE(your_data = belderbos_dataframe,
                          meta_data = belderbos_metadata,
                          threshold = 0)

# Make Belderbos celltype label consistent with other datasets
colData(belderbos_SE)$celltype <- gsub("G","Gr",colData(belderbos_SE)$celltype)
colData(belderbos_SE)$SAMPLENAME <- gsub("G","Gr",colData(belderbos_SE)$SAMPLENAME)
colnames(belderbos_SE) <- gsub("G","Gr",colnames(belderbos_SE))

system.file("sample_data/SixE_etal/WAS5_reads.txt", package = "barcodetrackR") %>%
  read.delim(row.names = 1) -> six_dataframe

system.file("sample_data/SixE_etal/WAS5_metadata.txt", package = "barcodetrackR") %>%
  read.delim() -> six_metadata

six_SE <- create_SE(your_data = six_dataframe,
                    meta_data = six_metadata,
                    threshold = 0.0001)

system.file("sample_data/EspinozaDA_etal/zl34_count_matrix.txt", package = "barcodetrackR") %>%
  read.delim(row.names = 1) -> esp_dataframe

system.file("sample_data/EspinozaDA_etal/zl34_metadata.txt", package = "barcodetrackR") %>%
  read.delim() -> esp_metadata

esp_SE <- create_SE(your_data = esp_dataframe,
                    meta_data = esp_metadata,
                    threshold = 0.0001)

system.file("sample_data/ElderA_etal/L4951_count_matrix.txt", package = "barcodetrackR") %>%
  read.delim(row.names = 1) -> eld_dataframe

system.file("sample_data/ElderA_etal/L4951_metadata.txt", package = "barcodetrackR") %>%
  read.delim() -> eld_metadata

eld_SE <- create_SE(your_data = eld_dataframe,
                    meta_data = eld_metadata,
                    threshold = 0.0001)

system.file("sample_data/ClarkeEL_etal/tcr_reads_p00007.txt", package = "barcodetrackR") %>%
  read.delim(row.names = 1) -> clarke_dataframe

system.file("sample_data/ClarkeEL_etal/tcr_metadata_p00007.txt", package = "barcodetrackR") %>%
  read.delim() -> clarke_metadata

clarke_SE <- create_SE(your_data = clarke_dataframe,
                       meta_data = clarke_metadata,
                       threshold = 0)

# Fix error in sample names of clarke SE
clarke_SE@colData@rownames <- clarke_metadata$SAMPLENAME

# Change nmonths to months to make it consistent with naming convention of other datasets
colData(clarke_SE)$months <- colData(clarke_SE)$nmonths

```

## Set results directory

```{r set results dir, eval = TRUE}
#results_dir <- "/Users/mortlockrd/Desktop/barcodetrackR manuscript/figures"
# results_dir <- "~/Documents/dunbar_lab/projects/diego_barcodetrackR/manuscript_figures/"
results_dir <- "/Users/mortlockrd/Desktop/barcodetrackR manuscript/figures v9"
```

## Fig S1: Influence of normalization and thresholding

```{r supp figure normalization and thresholding, eval = TRUE, warning = FALSE, message = FALSE}

p1 <- stat_hist(belderbos_SE[,5],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Belderbos et al B cells 22wk timepoint")

p2 <- stat_hist(belderbos_SE[,5],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(belderbos_SE)))


belderbos_SE2 <- create_SE(your_data = belderbos_dataframe,
                          meta_data = belderbos_metadata,
                          threshold = 0.0001)

p3 <- stat_hist(belderbos_SE2[,5],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Threshold of 0.01%")


p4 <- stat_hist(belderbos_SE2[,5],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(belderbos_SE2)))

belderbos_SE3 <- create_SE(your_data = belderbos_dataframe,
                          meta_data = belderbos_metadata,
                          threshold = 0.001)

p5 <- stat_hist(belderbos_SE3[,5],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Threshold of 0.1%")


p6 <- stat_hist(belderbos_SE3[,5],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(belderbos_SE3)))

six_SE <- create_SE(your_data = six_dataframe,
                    meta_data = six_metadata,
                    threshold = 0)

p7 <- stat_hist(six_SE[,5],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Six et al Monocytes 13m timepoint")


p8 <- stat_hist(six_SE[,5],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(six_SE)))

six_SE2 <- create_SE(your_data = six_dataframe,
                    meta_data = six_metadata,
                    threshold = 0.0001)

p9 <- stat_hist(six_SE2[,5],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Threshold of 0.01%")


p10 <- stat_hist(six_SE2[,5],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(six_SE2)))

six_SE3 <- create_SE(your_data = six_dataframe,
                    meta_data = six_metadata,
                    threshold = 0.001)

p11 <- stat_hist(six_SE3[,5],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Threshold of 0.1%")


p12 <- stat_hist(six_SE3[,5],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(six_SE3)))

eld_SE <- create_SE(your_data = eld_dataframe,
                    meta_data = eld_metadata,
                    threshold = 0)

p13 <- stat_hist(eld_SE[,1],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Elder et al Mouse AE3 spleen")

p14 <- stat_hist(eld_SE[,1],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(eld_SE)))

eld_SE2 <- create_SE(your_data = eld_dataframe,
                    meta_data = eld_metadata,
                    threshold = 0.0001)

p15 <- stat_hist(eld_SE2[,1],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Threshold of 0.01%")

p16 <- stat_hist(eld_SE2[,1],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(eld_SE2)))

eld_SE3 <- create_SE(your_data = eld_dataframe,
                    meta_data = eld_metadata,
                    threshold = 0.001)

p17 <- stat_hist(eld_SE3[,1],
                data_choice = "assay stats",
                assay_choice = "percentages",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = "Threshold of 0.1%")

p18 <- stat_hist(eld_SE3[,1],
                data_choice = "assay stats",
                assay_choice = "logs",
                y_log_axis = TRUE,
                n_bins = 20,
                your_title = paste0("Number of unique barcodes: ",nrow(eld_SE3)))

pdf(file = file.path(results_dir, "Figure_S1.pdf"), width = 24, height = 12)
cowplot::plot_grid(p1,p2,p3,p4,p5,p6,
                   p7,p8,p9,p10,p11,p12,
                   p13,p14,p15,p16,p17,p18,
                   ncol = 6, nrow = 3,
                   labels = c("A","","B","","C","",
                              "D","","E","","F","",
                              "G","","H","","I",""), align = "h", axis = "b")
dev.off()
```


## Figure 2: Global Clonal Distribution

```{r figure 2, eval = TRUE}

fig2_text_size <- 6 

six_cor_plot_sample_selection <- colData(six_SE) %>% as.data.frame() %>% dplyr::mutate(celltype = factor(celltype, levels = c("T", "B", "Gr", "Mo", "NK"))) %>% dplyr::arrange(celltype, months) %>% dplyr::pull(SAMPLENAME)

six_cor_plot_simple_names <- c("m13_T","m36_T","m43_T","m55_T",
                               "m13_B","m36_B","m43_B","m55_B",
                               "m13_Gr","m36_Gr","m43_Gr","m55_Gr",
                               "m13_Mo","m36_Mo","m43_Mo","m55_Mo",
                               "m13_NK","m36_NK","m43_NK","m55_NK")

six_cor_plot <- cor_plot(six_SE[,six_cor_plot_sample_selection],
                         method_corr = "pearson",
                         plot_type = "color",
                         label_size = fig2_text_size,
                         plot_labels = six_cor_plot_simple_names)+
  ggplot2::theme(legend.key.width=ggplot2::unit(0.2, "cm"))

bel_cor_plot_sample_selection <- colData(belderbos_SE) %>%
  as.data.frame() %>%
  dplyr::filter(weeks == "sac") %>%
  dplyr::arrange(organ, celltype) %>%
  dplyr::pull(SAMPLENAME) 

bel_cor_plot <- cor_plot(belderbos_SE[,bel_cor_plot_sample_selection],
                         method_corr = "pearson",
                         plot_type = "color",
                         label_size = fig2_text_size,
                         plot_labels = gsub("sac_","",colData(belderbos_SE[,bel_cor_plot_sample_selection])$SAMPLENAME))+
  ggplot2::theme(legend.key.width=ggplot2::unit(0.2, "cm"))

eld_cor_plot_sample_selection <- colData(eld_SE) %>% as.data.frame() %>% dplyr::filter(mouse %in% c("AE12", "AE32", "AE88", "AE89", "AE90")) %>% dplyr::pull(SAMPLENAME)
eld_cor_plot_labels <- colData(eld_SE) %>% as.data.frame() %>% dplyr::filter(mouse %in% c("AE12", "AE32", "AE88", "AE89", "AE90")) %>% dplyr::pull(short_name)

eld_cor_plot <- cor_plot(eld_SE[,eld_cor_plot_sample_selection],
                         method_corr = "pearson",
                         plot_type = "color",
                         label_size = fig2_text_size,
                         plot_labels = eld_cor_plot_labels)+
  ggplot2::theme(legend.key.width=ggplot2::unit(0.2, "cm"))

six_mds <- mds_plot(six_SE,
                    group_by = "celltype",
                    method_dist = "bray",
                    assay = "percentages",
                    point_size = 2,
                    text_size = fig2_text_size)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))+
  ggrepel::geom_text_repel(aes(label = months), size = fig2_text_size*0.5, show.legend = FALSE)+
  ggplot2::theme(plot.margin = ggplot2::unit(c(1.5,0,1,0), "lines")) 

bel_mds <- mds_plot(belderbos_SE[,bel_cor_plot_sample_selection],
                    group_by = "organ",
                    method_dist = "bray",
                    assay = "percentages",
                    point_size = 2,
                    text_size = fig2_text_size)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))+
  ggrepel::geom_text_repel(aes(label = celltype), size = fig2_text_size*0.5, show.legend = FALSE)+
  ggplot2::theme(plot.margin = ggplot2::unit(c(1.5,0,1,0), "lines")) 

# Shorten organ namee
colData(eld_SE)$organ <- gsub("part ","",colData(eld_SE)$organ)

eld_mds <- mds_plot(eld_SE[,eld_cor_plot_sample_selection],
                    group_by = "mouse",
                    method_dist = "bray",
                    assay = "percentages",
                    point_size = 2,
                    text_size = fig2_text_size)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))+
  ggrepel::geom_text_repel(aes(label = organ), size = fig2_text_size*0.5, show.legend = FALSE)+
  ggplot2::theme(plot.margin = ggplot2::unit(c(1.5,0,1,0), "lines")) 


pdf(file = file.path(results_dir, "Figure_2.pdf"), width = 8.5, height = 11)
cowplot::plot_grid(six_cor_plot, six_mds,
                   bel_cor_plot, bel_mds,
                   eld_cor_plot, eld_mds,
                   ncol = 2, nrow = 3,
                   labels = LETTERS[1:6], align = "h", axis = "b")
dev.off()
```

## Figure 3: Clone number and diversity

```{r figure 3, eval = TRUE}

fig3_ptsz <- 1.5
fig3_txsz <- 8

six_cc <- clonal_count(six_SE,
                       plot_over = "months",
                       group_by = "celltype",
                       cumulative = FALSE,
                       point_size = fig3_ptsz,
                       line_size = 1,
                       text_size = fig3_txsz)+
  ggplot2::theme(legend.position = "none")

six_cd <- clonal_diversity(six_SE,
                           plot_over = "months",
                           group_by = "celltype",
                           index_type = "shannon",
                           point_size = fig3_ptsz,
                           line_size = 1,
                           text_size = fig3_txsz)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))

bel_cc <- clonal_count(belderbos_SE[,c(1:6, 13:16)],
                       plot_over = "weeks",
                       group_by = "celltype",
                       cumulative = FALSE,
                       point_size = fig3_ptsz,
                       line_size = 1,
                       text_size = fig3_txsz)+
  ggplot2::theme(legend.position = "none")

bel_cd <- clonal_diversity(belderbos_SE[,c(1:6, 13:16)],
                           plot_over = "weeks",
                           group_by = "celltype",
                           index_type = "shannon",
                           point_size = fig3_ptsz,
                           line_size = 1,
                           text_size = fig3_txsz)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))

# Make espinoza cell type naming consistent with Six
colData(esp_SE)$celltype <- gsub("Mono","Mo",colData(esp_SE)$celltype)

# Name metadata days not day
colnames(colData(esp_SE))[2] <- "days"

esp_cc <- clonal_count(esp_SE[,colData(esp_SE)$celltype != "EOS"],
                       plot_over = "days",
                       group_by = "celltype",
                       cumulative = FALSE,
                       point_size = fig3_ptsz,
                       line_size = 1,
                       text_size = fig3_txsz)+
  ggplot2::theme(legend.position = "none")

esp_cd <- clonal_diversity(esp_SE[,colData(esp_SE)$celltype != "EOS"],
                           plot_over = "days",
                           group_by = "celltype",
                           index_type = "shannon",
                           point_size = fig3_ptsz,
                           line_size = 1,
                           text_size = fig3_txsz)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))+
  ggplot2::theme(plot.margin = ggplot2::unit(c(1.5,0,0,0), "lines")) # Will be reflected in all of them

pdf(file = file.path(results_dir, "Figure_3.pdf"), width = 7.5, height = 8)
cowplot::plot_grid(six_cc, six_cd, 
                   bel_cc, bel_cd, 
                   esp_cc, esp_cd,
                   ncol = 2, nrow = 3, align = "h",
                   rel_widths = c(0.86,1),
                   labels = LETTERS[1:6])
dev.off()

```

## Figure S2: Cumulative clone number

```{r supplemental figure cumulative clone count, eval = TRUE}

six_cumc <- clonal_count(six_SE,
                       plot_over = "months",
                       group_by = "celltype",
                       cumulative = TRUE,
                       point_size = fig3_ptsz,
                       line_size = 1,
                       text_size = fig3_txsz)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))

bel_cumc <- clonal_count(belderbos_SE[,c(1:6, 13:16)],
                       plot_over = "weeks",
                       group_by = "celltype",
                       cumulative = TRUE,
                       point_size = fig3_ptsz,
                       line_size = 1,
                       text_size = fig3_txsz)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))

esp_cumc <- clonal_count(esp_SE[,colData(esp_SE)$celltype != "EOS"],
                       plot_over = "days",
                       group_by = "celltype",
                       cumulative = TRUE,
                       point_size = fig3_ptsz,
                       line_size = 1,
                       text_size = fig3_txsz)+
  ggplot2::theme(legend.key.height = ggplot2::unit(0.25, "cm"))+
  ggplot2::theme(plot.margin = ggplot2::unit(c(1.5,0,0,0), "lines")) # Will be reflected in all of them


pdf(file = file.path(results_dir, "Figure_S2.pdf"), width = 10, height = 3)
cowplot::plot_grid(six_cumc,bel_cumc, esp_cumc,
                   ncol = 3, nrow = 1, align = "h",
                   labels = LETTERS[1:3])
dev.off()

```


## Figure 4: Tracking individual clones

```{r figure 4, echo=TRUE, warning = FALSE, eval = TRUE}

fig3_txsz <- 8

wu_CD16_NK_order <- c("ZJ31_3m_NK_CD56n_CD16p", "ZJ31_4m_NK_CD56n_CD16p",
                      "ZJ31_6m_NK_CD56n_CD16p", "ZJ31_8.5m_NK_CD56n_CD16p",
                      "ZJ31_9.5m_NK_CD56n_CD16p", "ZJ31_12m_NK_CD56n_CD16p",
                      "ZJ31_17.5m_NK_CD56n_CD16p", "ZJ31_20m_NK_CD56n_CD16p")

wu_hm_bh <- barcode_binary_heatmap(your_SE = wu_SE[,wu_CD16_NK_order],
                                   label_size = fig3_txsz,
                                   plot_labels = c("3m NK 16", "4m NK 16", "6m NK 16",
                                                   "8.5m NK 16", "9.5m NK 16", "12m NK 16",
                                                   "17.5m NK 16", "20m NK 16"),
                                   your_title = " ")
wu_hm_tc <- barcode_ggheatmap(your_SE = wu_SE[,wu_CD16_NK_order],
                              n_clones = 10,
                              cellnote_assay = "stars",
                              cellnote_size = 4,
                              label_size = fig3_txsz,
                              plot_labels = c("3m NK 16", "4m NK 16", "6m NK 16",
                                              "8.5m NK 16", "9.5m NK 16", "12m NK 16",
                                              "17.5m NK 16", "20m NK 16"),
                              your_title = " ")
wu_hm_st <- barcode_ggheatmap_stat(your_SE = wu_SE[,wu_CD16_NK_order],
                                   sample_size = rep(40000,length(wu_CD16_NK_order)),
                                   stat_test = "chi-squared",
                                   stat_option = "subsequent",
                                   p_threshold = 0.05,
                                   n_clones = 10,
                                   cellnote_assay = "stars",
                                   cellnote_size = 4,
                                   label_size = fig3_txsz,
                                   plot_labels = c("3m NK 16", "4m NK 16", "6m NK 16",
                                                   "8.5m NK 16", "9.5m NK 16", "12m NK 16",
                                                   "17.5m NK 16", "20m NK 16"),
                                   your_title = " ")

pdf(file = file.path(results_dir, "Figure_4.pdf"), width = 12, height = 5)
cowplot::plot_grid(wu_hm_bh, wu_hm_tc, wu_hm_st, ncol = 3, labels = c("A", "B", "C"))
dev.off()

```


## Figure 5: Clonal bias comparing Gr and T

```{r figure 5, eval = TRUE}

fig5_txsz <- 8

# Ridge plots
six_ridge <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "Gr",
                        bias_2 = "T",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = fig5_txsz)

ko_ridge <- bias_ridge_plot(your_SE = koelle_SE,
                       split_bias_on = "celltype",
                       bias_1 = "Gr",
                       bias_2 = "T",
                       split_bias_over = "months",
                       bias_over = c(4.5,21,30,38),
                       weighted = TRUE,
                       add_dots = TRUE,
                       text_size = fig5_txsz)

# Circos plots

# Make koelle naming consistent with six
colData(koelle_SE)$celltype <- gsub("Mono",  "Mo", colData(koelle_SE)$celltype)

# When saving circos plots to variables, I have to use this hacky way of doing it :/
ko_circos_selection <- c("ZH33_38m_Gr_Ficoll","ZH33_38m_T","ZH33_38m_Mono")
ko_circos <- function() chord_diagram(koelle_SE[,ko_circos_selection],
                                    weighted = TRUE,
                                    plot_label = "celltype",
                                    text_size = fig5_txsz)
six_circos_selection <- c("m55_GRANULOCYTES","m55_TCELLS","m55_MONOCYTES")
six_circos <- function() chord_diagram(six_SE[,six_circos_selection],
                                     weighted = TRUE,
                                     plot_label = "celltype",
                                     text_size = fig5_txsz)

pdf(file = file.path(results_dir, "Figure_5.pdf"), width = 8, height = 8)
cowplot::plot_grid(six_ridge, six_circos, ko_ridge, ko_circos,
                   ncol = 2, nrow = 2, align = "h",
                   labels = c("A", "B", "C", "D"))
dev.off()

```

## Figure S3: Further Clonal Bias from Six et al

```{r figure S3, eval = TRUE}

figS3_txsz <- 8

# Ridge plots
six_ridge2 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "Gr",
                        bias_2 = "Mo",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

six_ridge3 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "Gr",
                        bias_2 = "B",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

six_ridge4 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "Gr",
                        bias_2 = "NK",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

six_ridge5 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "Mo",
                        bias_2 = "T",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

six_ridge6 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "Mo",
                        bias_2 = "B",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

six_ridge7 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "Mo",
                        bias_2 = "NK",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

six_ridge8 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "T",
                        bias_2 = "B",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

six_ridge9 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "T",
                        bias_2 = "NK",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

six_ridge10 <- bias_ridge_plot(your_SE = six_SE,
                        split_bias_on = "celltype",
                        bias_1 = "B",
                        bias_2 = "NK",
                        split_bias_over = "months",
                        weighted = TRUE,
                        add_dots = TRUE,
                        text_size = figS3_txsz)

# Chord diagrams
six_circos_selection2 <- c("m55_GRANULOCYTES","m55_BCELLS","m55_MONOCYTES")
six_circos2 <- function() chord_diagram(six_SE[,six_circos_selection2],
                                     weighted = TRUE,
                                     plot_label = "celltype",
                                     text_size = figS3_txsz)

six_circos_selection3 <- c("m55_GRANULOCYTES","m55_NKCELLS","m55_MONOCYTES")
six_circos3 <- function() chord_diagram(six_SE[,six_circos_selection3],
                                     weighted = TRUE,
                                     plot_label = "celltype",
                                     text_size = figS3_txsz)

six_circos_selection4 <- c("m55_TCELLS","m55_BCELLS","m55_NKCELLS")
six_circos4 <- function() chord_diagram(six_SE[,six_circos_selection4],
                                     weighted = TRUE,
                                     plot_label = "celltype",
                                     text_size = figS3_txsz)

pdf(file = file.path(results_dir, "Figure_S3.pdf"), width = 11, height = 8)
cowplot::plot_grid(six_ridge2, six_ridge3, six_ridge4, six_circos2,
                   six_ridge5, six_ridge6, six_ridge7, six_circos3,
                   six_ridge8, six_ridge9, six_ridge10, six_circos4,
                   ncol = 4, nrow = 3, align = "h",
                   labels = LETTERS[1:12])
dev.off()

```


## Figure S4: Further Clonal Bias from Koelle et al

```{r figure S4, eval = TRUE}

figS4_txsz <- 8

# Ridge plots
ko_ridge2 <- bias_ridge_plot(your_SE = koelle_SE,
                       split_bias_on = "celltype",
                       bias_1 = "Gr",
                       bias_2 = "Mo",
                       split_bias_over = "months",
                       bias_over = c(4.5,21,30,38),
                       weighted = TRUE,
                       add_dots = TRUE,
                       text_size = figS4_txsz)

ko_ridge3 <- bias_ridge_plot(your_SE = koelle_SE,
                       split_bias_on = "celltype",
                       bias_1 = "Mo",
                       bias_2 = "T",
                       split_bias_over = "months",
                       bias_over = c(4.5,21,30,38),
                       weighted = TRUE,
                       add_dots = TRUE,
                       text_size = figS4_txsz)

pdf(file = file.path(results_dir, "Figure_S4.pdf"), width = 8, height = 4)
cowplot::plot_grid(ko_ridge2, ko_ridge3,
                   ncol = 2, nrow = 1, align = "h",
                   labels = LETTERS[1:2])
dev.off()

```


## Figure S5: Clone number and diversity of Clarke et al TCR data

```{r figure S4, eval = TRUE}

figS5_text_size <- 6 

clarke_cc <- clonal_count(clarke_SE,
                              plot_over = "months",
                              group_by = "SAMPLENAME",
                              cumulative = FALSE,
                              point_size = 2,
                              line_size = 1,
                              text_size = figS5_text_size)+
  ggplot2::theme(legend.position = "none")

clarke_cd <- clonal_diversity(clarke_SE,
                              plot_over = "months",
                              group_by = "SAMPLENAME",
                              index_type = "shannon",
                              point_size = 2,
                              line_size = 1,
                              text_size = figS5_text_size)+
  ggplot2::theme(legend.position = "none")+
  ggplot2::theme(plot.margin = ggplot2::unit(c(1.5,0,0,0), "lines")) # Will be reflected in all of them

pdf(file = file.path(results_dir, "Figure_S5.pdf"), width = 7, height = 8/3)
cowplot::plot_grid(clarke_cc, clarke_cd,
                   ncol = 2, nrow = 1, align = "h",
                   labels = LETTERS[1:2])
dev.off()

```
